---
output:
  pdf_document: default
  html_document: default
  word_document: default
---
##    PACKAGES USED
```{r}
mozart <- c("readxl","dplyr","tidyverse","ggplot2","readr","forecast","tseries","astsa","dplyr","tidyverse")
lapply(mozart, library, character.only = TRUE)
```

##    GET DATA
```{r}
fuel <- read_excel("~/Copy of petrol and diesel time series data.xlsx")
View(fuel)
gdp <- read.csv("export.csv")
head(fuel)
petrol <- fuel$AVERAGE.PETROL
diesel <- fuel$AVERAGE.DIESEL
head(gdp)
head(petrol)
head(diesel)
```


##    CHECK FOR OUTLIERS
GDP has outliers at the values 1838.2100 1816.5471 1708.0014 1572.3457 1410.5320 1336.8833 1315.8022 1210.3854 1136.8699 971.6330 951.6878
```{r}
boxplot(diesel)
boxplot(diesel)$out
boxplot(gdp$GDP.Per.Capita)$out
```

##    REMOVING OUTLIERS

###   GET THE ROWS THET ARE IN
```{r}
outliers <- boxplot(gdp$GDP.Per.Capita, plot = FALSE)$out
boxplot(outliers)
gdp.out <- gdp[which(gdp$GDP.Per.Capita %in% outliers),]
gdp.out <- gdp.out
gdp.out
```

##    PREPARING THE TIMESERIES
```{r}
gdp.ts <-  ts(gdp$GDP.Per.Capita, start = 1960, frequency = 1)
gdp.ts
autoplot(gdp.ts)
petrol.ts <- ts(petrol, frequency = 12, start = c(2010,12))
petrol.ts
autoplot(petrol.ts)
diesel.ts <- ts(diesel, frequency = 12, start = c(2010,12))
autoplot(diesel.ts)
```

##    STATIONERITY

All the three were non stationery
```{r}
diesel.st <- adf.test(diesel.ts)
diesel.st
gdp.st <- adf.test(gdp.ts)
gdp.st
petrol.st <-adf.test(petrol.ts)
petrol.st


kpss.test(diesel.ts)
kpss.test(petrol.ts)
kpss.test(gdp.ts)


```
p-value>0.1 indicates stationary


##    GETTING THE TRENDS AND SEASONALITY
```{r}
autoplot(decompose(diesel.ts, type = "multiplicative"))
autoplot(decompose(diesel.ts))
autoplot(decompose(petrol.ts))
autoplot(decompose(petrol.ts, type = "multiplicative"))
GGally::ggpairs(diesel.ts)
GGally::ggpairs(petrol.ts)
GGally::ggpairs(gdp.ts)
petrol.ts
gdp.ts
diesel.ts

```

##    MOVING AVERAGES
```{r}
gdp.ma <- TTR::SMA(gdp$GDP.Per.Capita, n = 4)
gdp.ma
petrol.ma <- TTR::SMA(petrol, n = 4)
diesel.ma <- TTR::SMA(diesel, n = 4)
plot(gdp.ma, type = "l", col = "green")
plot(petrol.ma, type = "l", col = "blue")
lines(diesel.ma, type = "l")
```

##    AUTOCORELATION FUNCTION
due to non stationerity we differenciate

PACF GIVES THE AR
ACF GIVES  THE MA
```{r}
diesel.diff <- diff(diesel.ts)
diesel.diff
gdp.diff <- diff(gdp.ts)
petrol.diff <- diff(petrol.ts)

gdp.acf <- acf(gdp.diff)#1,2,3,6,10,11
gdp.pacf <- pacf(gdp.diff)#1,4,13

petrol.acf <- acf(petrol.diff)#0,1
petrol.pacf <- pacf(petrol.diff)#0,1

diesel.diff2 <- diff(diesel.ts,differences = 2)#lag = 3, differences = 2)

diesel.acf <- acf(diesel.diff2,lag.max = 10)#0,1
diesel.pacf <- pacf(diesel.diff2)#1,2,3
```



##    FITTING AR MODEL
gdp has AR of AR(1) and AR(4)

gdp has MA of MA(1), MA(2) and MA(3)

check the likelihood for the best model

best model for gdp is ARIMA(4,1,1) it has the lowest log likelihood

petrol has a AR of AR(0) ans MA of MA(0)

ARIMA(0,1,0)

diesel has best ARIMA at ARIMA(1,2,0)
```{r}
#1,2,3,6,10,11
#1,4,13

gdp.arima <- arima(gdp.diff,order = c(1,1,1))
gdp.arima1 <- arima(gdp.diff,order = c(1,1,2))
gdp.arima2 <- arima(gdp.diff,order = c(1,1,3))
gdp.arima3 <- arima(gdp.diff,order = c(1,1,6))
gdp.arima4 <- arima(gdp.diff,order = c(1,1,10))
gdp.arima5 <- arima(gdp.diff,order = c(1,1,11))
```


```{r}
gdp.arima6 <- arima(gdp.diff,order = c(4,1,1))
gdp.arima7 <- arima(gdp.diff,order = c(4,1,2))

#gdp.arima8 <- arima(gdp.diff,order = c(4,1,3))
gdp.arima9 <- arima(gdp.diff,order = c(4,1,6))
gdp.arima10 <- arima(gdp.diff,order = c(4,1,10))

#gdp.arima11 <- arima(gdp.diff,order = c(4,1,11))

#gdp.arima12 <- arima(gdp.diff,order = c(13,1,1))
gdp.arima13 <- arima(gdp.diff,order = c(13,1,2))
gdp.arima14 <- arima(gdp.diff,order = c(13,1,3))
gdp.arima15 <- arima(gdp.diff,order = c(13,1,6))
gdp.arima16 <- arima(gdp.diff,order = c(13,1,10))
gdp.arima17 <- arima(gdp.diff,order = c(13,1,11))
```


```{r}
gdp.arima 
gdp.arima1 
gdp.arima2 
gdp.arima3 
gdp.arima4 
gdp.arima5
```


```{r}
gdp.arima6 
gdp.arima7 
#gdp.arima8 
gdp.arima9 
gdp.arima10
#gdp.arima11
```


```{r}
#gdp.arima12
gdp.arima13
#gdp.arima14
gdp.arima15
gdp.arima16
gdp.arima17
```


```{r}
gdp.arima$loglik
gdp.arima1$loglik
gdp.arima2$loglik
gdp.arima3$loglik
gdp.arima4$loglik


petrol.arima <- arima(diesel.diff,order = c(0,1,0))
petrol.arima1 <- arima(diesel.diff,order = c(0,1,1))
petrol.arima2 <- arima(diesel.diff,order = c(1,1,0))
petrol.arima3 <- arima(diesel.diff,order = c(1,1,1))

petrol.arima
petrol.arima1
petrol.arima2
petrol.arima3



diesel.arima <- arima(diesel.diff,order = c(1,2,0))
diesel.arima1 <- arima(diesel.diff,order = c(1,2,2))
diesel.arima2 <- arima(diesel.diff,order = c(2,2,0))
diesel.arima3 <- arima(diesel.diff,order = c(2,2,1))
diesel.arima4 <- arima(diesel.diff,order = c(3,2,0))
diesel.arima5 <- arima(diesel.diff,order = c(3,2,1))
diesel.arima
diesel.arima1
diesel.arima2
diesel.arima3
diesel.arima4
diesel.arima5
```

##    TEST FOR  WHITENOISE

ARE THE RESIDUALS  OF ARIMA FOLLOWING THE WN PROCESS
From the ACF and PACF plots, we do not observe any auto-regressive nature in our residuals. The Ljung Box test, further confirms our belief that the residuals from the best fit model is not autocorrelated as p-value significantly greater than 0.05.
```{r}
LB_test <- Box.test(diesel.arima$residuals, type='Ljung-Box')
print(LB_test)
LB_test <- Box.test(petrol.arima$residuals, type='Ljung-Box')
print(LB_test)
LB_test <- Box.test(gdp.arima$residuals, type='Ljung-Box')
print(LB_test)

forecast::checkresiduals(petrol.arima3)
forecast::checkresiduals(diesel.arima1)
forecast::checkresiduals(gdp.arima17)

#forecast::checkresiduals(petrol.arima1)
#forecast::checkresiduals(diesel.arima1)

acf(diesel.arima1$residuals)
acf(petrol.arima3$residuals)
acf(gdp.arima17$residuals)

pacf(diesel.arima1$residuals)
pacf(gdp.arima17$residuals)
pacf(petrol.arima3$residuals)


```

##    FORECAST
```{r}
diesel.arima1 %>%
  forecast(h = 5) %>%
  autoplot()

gdp.arima17 %>%
  forecast(h = 5) %>%
  autoplot()

petrol.arima3 %>%
  forecast(h = 5) %>%
  autoplot()


diesel.ts %>%
  forecast(h = 10) %>%
  autoplot()

gdp.ts %>%
  forecast(h = 10) %>%
  autoplot()

petrol.ts %>%
  forecast(h = 10) %>%
  autoplot()
```
##analysis

```{r}
library(readxl)
generated <- read_excel("C:/Users/Meshack/Desktop/generated.xlsx", 
    sheet = "Sheet1")
View(generated)
pol <- generated
pol

library(dplyr)

New_pol <- pol %>% select(-Time, -Rain_season, -Wind_Speed, -Abudance, -Insect_Family, -Insect_Order, -Insect_Species, -Insect_act, -collector_name)

View(New_pol)

model <- lm(New_pol$Insect_ID ~ New_pol$Temp + New_pol$Cloud_Cover)

model

insect <- (-0.5943) + (0.1732 * 25) - (-0.0204 * 25)

insect
```